---
title: "Product Demand Forecast (hts)"
author: "Microsoft"
output: 
    rmarkdown::html_vignette:
        toc: true

vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, echo=FALSE}

knitr::opts_chunk$set(fig.width = 6,
                      fig.height = 4,
                      fig.align ='center',
                      dev = "png")

```

# Introduction

In this documentation, we are trying to build a quartly by category demand forecast model on product booking quantity data using the package 'hts'. We implement the advanced analytics process with R in the principle of step-by-step. 

## Product booking quantity data

```{r, message=FALSE}
## Load the required packages into the R session.

library(dplyr)        # Missing value treatment with Filter.
library(zoo)          # Missing value treatment with locf method.
library(hts)          # Hierarchical time series forecasting.
library(fpp)          # Time Series cross validation.
library(foreach)      # Run for loop in parallel.
library(RevoScaleR)   # Enable out-of-memory computation using rxExec in R.
```

Let's load the data and take a brief glimpse of it. 

```{r, message=FALSE}
## Set directory

setwd("C:/Users/zhouf/Documents/GitHub/acceleratoRs/ProductDemandForecast")

wd <- getwd()
```

```{r, message=FALSE, warning=FALSE, error=FALSE}
## Read data

dspath <- file.path(wd, "Data", "productQuantity.csv")

data <- read.table(dspath, header=TRUE, sep=",")

dim(data) # 56 11

str(data)

head(data)
```

## Handling missing value

To clean the data, we remove some of columns with too many NAs (>10%) and replace other missing values with last-observation-carried-forward method. 

```{r, message=FALSE, warning=FALSE, error=FALSE}
## Clean missing value

# Remove columns with only NA

data <- Filter(function(x)!all(is.na(x)), data)
dim(data) # 56 10

# Remove columns with too many NAs 

data <- Filter(function(x)!(sum(is.na(x))/length(x)) > 0.1, data)
dim(data) # 56 10

# Remove rows with only missing values

missingRows <- apply(data, 1, function(x) all(is.na(x)))
data <- data[!missingRows, ]
dim(data)

# Replace NA values with last-observation-carried-forward method

data[,3:10] <- lapply(data[,3:10], function(x) { 
  x <- na.locf(x)
  x
})

head(data)
```

## Model building

We aim to build a time series model to fit the whole product booking quantity data. 

### Hierachical time series model

Hierarchical time series forecasting methods can be implemented in the hts package (Hyndman et al. 2014) for R (R Core Team 2013).

We can use hts function to create a hierarchical time series. The required inputs are the bottom-level time series, and information about the hierarchical structure. Here, we try two hierarchical structure: first, each bottom series belong to a group; second, the series for product 1 and 2 belong to category A, series for product 3-8 belong to the category B.

Let's have a look at this 2-level hierarchical time series. 

```{r, message=FALSE, warning=FALSE, error=FALSE}
## Hierarchical time series modeling 

# Construct time series matrix

bts <- ts(data[, 3:10], frequency=4, start=c(1998, 1))

head(bts)

# Create 2-level hierarchical time series

demandgts <- hts(bts, nodes=list(2, c(2, 6)))

# Rename the nodes of the hierarchy

demandgts$labels$`Level 1` <- c("Category.A", "Category.B")
demandgts$labels$`Level 2` <- c("Product.1", "Product.2", "Product.3", "Product.4", 
                                "Product.5", "Product.6", "Product.7", "Product.8")
names(demandgts$labels) <- c("Total", "Category", "Product")

head(demandgts)
print(demandgts)

# Aggregate the hierarchy time series

alldemand <- aggts(demandgts) 

head(alldemand)

plot(demandgts, levels=c(0, 1, 2))

# Set up the training sample and testing sample

traingts <- window(demandgts, start=c(1998, 1), end=c(2009, 4))
testgts <- window(demandgts, start=c(2010, 1), end=c(2011, 4))

demandfcast <- forecast(object  = traingts, 
                        h       = 8, 
                        method  = "bu", 
                        fmethod = "arima",
                        keep.fitted = TRUE)

# Explore and visualize the forecasted hts

aggts(demandfcast)

summary(demandfcast)

plot(demandfcast)

# Calculate perforemance metrics

accuracy.gts(demandfcast, testgts, levels=c(0, 1, 2))
```

### Cross-validation for hierarchical time series

Next, we use cross validation to compare performance of different forecast methods for hierarchical time series model.

```{r, message=FALSE, warning=FALSE, error=FALSE}
## Time series cross validation

# Define function that runs hierarchical forecast on training data set
# and returns forecast on test data set

forecast_hts <- function (traindata, testdata, htsmethod, tsmethod) {
  
  # Forecasting horizon
  
  horiz =  dim(aggts(testdata))[1]

  # Run hierarchical forecast
  
  hts_fcast <- forecast(object  = traindata, 
                        h       = horiz,
                        method  = htsmethod,
                        fmethod = tsmethod,
                        keep.fitted = TRUE)
  
  # Return forecasted ts
  
  hts_fcast
  
  # Return evaluation metrics at the top level
  #fcast_acc <- accuracy.gts(hts_fcast, test = testdata, levels = 0)
}

# Set parameters for cross validation

k <- 48 
n <- dim(demandgts$bts)[1]
st <- tsp(demandgts$bts)[1]+(k-2)/4

# Define parameter space for hts

hts_method <- c("bu", "comb", "tdgsa")
ts_method <- c("arima", "ets", "rw")

param_space <- expand.grid(hts_method, 
                           ts_method, 
                           stringsAsFactors = FALSE)

colnames(param_space) <- c("hts_method", "ts_method")

# Define empty matric matrix

mape11 <- mape12 <- mape13 <- mape21 <- mape22 <- mape23 <- mape31 <- mape32 <- mape33 <- matrix(NA,n-k,4)

rxSetComputeContext(RxLocalParallel())

tscv_time <- system.time({
  
foreach(i=1:(n-k)) %dopar%
{
  datashort <- window(demandgts, end=st + i/4)
  datanext <- window(demandgts, start=st + (i+1)/4, end=st + (i+4)/4)
  
  # Run many distributed jobs

  rxResult <- rxExec(FUN=forecast_hts,  
                     traindata=datashort, 
                     testdata=datanext, 
                     htsmethod=rxElemArg(param_space$hts_method),  
                     tsmethod=rxElemArg(param_space$ts_method),
                     consoleOutput=TRUE,
                     packagesToLoad=c('hts'))
  
  horiz <- dim(datanext$bts)[1]
  
  mape11[i, 1:horiz] <- accuracy(aggts(rxResult[[1]]), aggts(datanext))[1, "MAPE"]
  mape12[i, 1:horiz] <- accuracy(aggts(rxResult[[2]]), aggts(datanext))[1, "MAPE"]
  mape13[i, 1:horiz] <- accuracy(aggts(rxResult[[3]]), aggts(datanext))[1, "MAPE"]
  mape21[i, 1:horiz] <- accuracy(aggts(rxResult[[4]]), aggts(datanext))[1, "MAPE"]
  mape22[i, 1:horiz] <- accuracy(aggts(rxResult[[5]]), aggts(datanext))[1, "MAPE"]
  mape23[i, 1:horiz] <- accuracy(aggts(rxResult[[6]]), aggts(datanext))[1, "MAPE"]
  mape31[i, 1:horiz] <- accuracy(aggts(rxResult[[7]]), aggts(datanext))[1, "MAPE"]
  mape32[i, 1:horiz] <- accuracy(aggts(rxResult[[8]]), aggts(datanext))[1, "MAPE"]
  mape33[i, 1:horiz] <- accuracy(aggts(rxResult[[9]]), aggts(datanext))[1, "MAPE"]
}
  
})

# Visualize overall mape vs horiz

plot(1:4, colMeans(mape11,na.rm=TRUE), type="o", col=2, 
     xlab="horizon", ylab="Overall MAPE", ylim=c(0, 25))
lines(1:4, colMeans(mape12,na.rm=TRUE), type="o",col=3)
lines(1:4, colMeans(mape13,na.rm=TRUE), type="o",col=4)
lines(1:4, colMeans(mape21,na.rm=TRUE), type="o",col=5)
lines(1:4, colMeans(mape22,na.rm=TRUE), type="o",col=6)
lines(1:4, colMeans(mape23,na.rm=TRUE), type="o",col=7)
lines(1:4, colMeans(mape31,na.rm=TRUE), type="o",col=8)
lines(1:4, colMeans(mape32,na.rm=TRUE), type="o",col=9)
lines(1:4, colMeans(mape33,na.rm=TRUE), type="o",col=10)
legend("topleft", 
       legend=c("bu-ARIMA", "bu-ETS", "bu-RW",
                "comb-ARIMA", "comb-ETS", "comb-RW",
                "tdgsa-ARIMA", "tdgsa-ETS", "tdgsa-RW"), 
       col=2:10, lty=1, cex=0.5)

# Show the run time of time series cross validation

print(tscv_time)

# Forecast the product demand of 2012 using the optimal parameters

opt_fcast <- forecast(object  = demandgts, 
                      h       = 4,
                      method  = "comb",
                      fmethod = "arima",
                      keep.fitted = TRUE)

# Explore and visualize the forecasted hts

aggts(opt_fcast)

summary(opt_fcast)

plot(opt_fcast)
```

