---
title: "Deploy a Credit Risk Model as a Web Service"
author: "Fang Zhou, Data Scientist, Microsoft"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.width = 8,
                      fig.height = 5,
                      fig.align='center',
                      dev = "png")
```

## 1 Introduction

The `mrsdeploy` package, delivered with Microsoft R Client and R Server, provides functions for:

**1** Establishing a remote session in a R console application for the purposes of executing code on that server

**2** Publishing and managing an R web service that is backed by the R code block or script you provided. 

Each feature can be used independently, but the greatest value is achieved when you can leverage both.

This document will walk through you how to deploy a credit risk model as a web service, using the `mrsdeploy` package.

It will start by modelling locally, then publish it as a web service, and then share it with other authenticated users for consumption, and finally manage and update the web service. 

## 2 Automated Credit Risk Model Deployment

### 2.1 Setup

We load the required R packages.

```{r, message=FALSE, warning=FALSE, error=FALSE}
## Setup

# Load the required packages into the R session.

library(rattle)       # Use normVarNames().
library(dplyr)        # Wrangling: tbl_df(), group_by(), print(), glimpse().
library(magrittr)     # Pipe operator %>% %<>% %T>% equals().
library(scales)       # Include commas in numbers.
library(MicrosoftML)  # Build models using Microsoft ML algortihms.
library(mrsdeploy)    # Publish an R model as a web service.
```

Then, the dataset transactionSimu, demographicSimu and processedSimu are ingested for demonstration. The processedSimu dataset was created by the data preprocessing steps in the data science accelerator for credit risk prediction.

```{r, message=FALSE, warning=FALSE, error=FALSE}
## Data Ingestion

# Identify the source location of the dataset.

#DATA <- "../../Data/"
#txn_fname <- file.path(DATA, "Raw/processedSimu.csv")

wd <- getwd()

dpath <- "../Data"
txn_fname   <- file.path(wd, dpath, "transactionSimu_v3.csv")
demog_fname <- file.path(wd, dpath, "demographicSimu_v3.csv")
data_fname <- file.path(wd, dpath, "processedSimu.csv")

# Ingest the dataset.

txn   <- read.csv(file=txn_fname)   %T>% 
   {dim(.) %>% comma() %>% cat("\n")}

demog <- read.csv(file=demog_fname) %T>% 
   {dim(.) %>% comma() %>% cat("\n")}

data <- read.csv(file=data_fname) %T>%
   {dim(.) %>% comma() %>% cat("\n")}

# A glimpse into the data.

glimpse(txn)
glimpse(demog)
glimpse(data)
```


### 2.2 Model Locally

Now, let's get started to build an R model based web service. 

First of all, we create a machine learning fast tree model on the dataset processedSimu by using the function `rxFastTrees()` from the `MicrosoftML` package. This model could be used to predict whether an account will default or to predict its probability of default, given some transaction statistics and demographic & bank account information as inputs.

```{r, message=FALSE, warning=FALSE, error=FALSE}
## Variable roles.

# Target variable

target <- "bad_flag"

# Note any identifier.

id <- c("account_id") %T>% print() 

# Note the available variables as model inputs.

vars <- setdiff(names(data), c(target, id))
```

```{r, message=FALSE, warning=FALSE, error=FALSE}
# Split Data

set.seed(42)

data <- data[order(runif(nrow(data))), ]

train <- sample(nrow(data), 0.70 * nrow(data))
test <- setdiff(seq_len(nrow(data)), train)
```

```{r, message=FALSE, warning=FALSE, error=FALSE}
# Prepare the formula

top_vars <- c("amount_6", "pur_6", "avg_pur_amt_6", "avg_interval_pur_6", "credit_limit", "age", "income", "sex", "education", "marital_status")

form <- as.formula(paste(target, paste(top_vars, collapse="+"), sep="~"))
form
```

```{r, message=FALSE, warning=FALSE, error=FALSE}
# Train model: rxFastTrees

model_rxtrees <- rxFastTrees(formula=form,
                             data=data[train, c(target, vars)],
                             type="binary",
                             numTrees=100,
                             numLeaves=20,
                             learningRate=0.2,
                             minSplit=10,
                             unbalancedSets=FALSE,
                             verbose=0)

model_rxtrees
```

```{r, message=FALSE, warning=FALSE, error=FALSE}
## Produce a function that can process the new input datasets.

creditRiskProcess <- function(txn, demog)
{ 
  require(rattle)
  require(dplyr)        
  require(magrittr) 
  
  # Import the datasets.
  
  txn <- rxImport(inData=txn, stringsAsFactors=TRUE)
  demog <- rxImport(inData=demog, stringsAsFactors=TRUE)
  
  # Normalize variable names.
  
  names(txn)   %<>% normVarNames() 
  names(demog) %<>% normVarNames() 
  
  # Transform date variable from factor to date
  
  txn$transaction_date <- as.Date(txn$transaction_date)
  
  # Normalize categoric variables.
  
  txn %>% 
    sapply(is.factor) %>%
    which() %>%
    names() ->
    txnc
  
  demog %>% 
    sapply(is.factor) %>%
    which() %>%
    names() ->
    demc
  
  txn[txnc]   %<>% sapply(normVarNames)
  demog[demc] %<>% sapply(normVarNames)
  
  # Define some functions to simply the calculation.

  # Calculate maximum interval of a time series.
  
  maxInterval <- function(ts)
  {
    if (length(na.omit(ts)) <= 1)
    {
      interval <- 0
    }
    else
    {
      interval <- as.numeric(max(difftime(c(ts[-1], NA), ts, units="day"), na.rm=TRUE))
    }
    return(interval)
  }
  
  # Calculate minimum interval of a time series.
  
  minInterval <- function(ts)
  {
    if (length(na.omit(ts)) <= 1)
    {
      interval <- 0
    }
    else
    {
      interval <- as.numeric(min(difftime(c(ts[-1], NA), ts, units="day"), na.rm=TRUE))
    }
    return(interval)
  }
  
  # Calculate average interval of a time series.
  
  avgInterval <- function(ts)
  {
    if (length(na.omit(ts)) <= 1)
    {
      interval <- 0
    } else
    {
      interval <- as.numeric(mean(difftime(c(ts[-1], NA), ts, units="day"), na.rm=TRUE))
    }
    return(interval)
  }
  
  # Calculate interval of time between enddate and the most recent transaction date.
  
  recentInterval <- function(startdate, enddate, ts)
  {
    if (length(na.omit(ts)) == 0)
    {
      interval <- as.numeric(difftime(enddate, startdate, units="day"))
    }
    else
    {
      interval <- as.numeric(difftime(enddate, max(ts, na.rm=TRUE), units="day"))
    }
    return(interval)
  }
  
  # Aggregate transaction data per account level.
  
  txn %>%
    filter(transaction_type == "p") %>%
    group_by(account_id) %>%
    arrange(account_id, transaction_date) %>%
    summarise(amount_6      = sum(transaction_amount_usd),
              pur_6         = n(),
              bank_6        = ifelse(pur_6 == 0, 0, sum(merchant_industry == "bank")/pur_6),
              entmnt_6      = ifelse(pur_6 == 0, 0, sum(merchant_industry == "entmnt")/pur_6),
              jewellery_6   = ifelse(pur_6 == 0, 0, sum(merchant_industry == "jewellery")/pur_6),
              medical_6     = ifelse(pur_6 == 0, 0, sum(merchant_industry == "medical")/pur_6),
              others_6      = ifelse(pur_6 == 0, 0, sum(merchant_industry == "others")/pur_6),
              petrol_6      = ifelse(pur_6 == 0, 0, sum(merchant_industry == "petrol")/pur_6),
              restaurant_6  = ifelse(pur_6 == 0, 0, sum(merchant_industry == "restaurant")/pur_6),
              supermarket_6 = ifelse(pur_6 == 0, 0, sum(merchant_industry == "supermkt")/pur_6),
              telecom_6     = ifelse(pur_6 == 0, 0, sum(merchant_industry == "telecom")/pur_6),
              travel_6      = ifelse(pur_6 == 0, 0, sum(merchant_industry == "travel")/pur_6),
              utility_6     = ifelse(pur_6 == 0, 0, sum(merchant_industry == "utility")/pur_6),
              avg_pur_amt_6 = ifelse(pur_6 == 0, 0, sum(transaction_amount_usd)/pur_6),
              max_pur_amt_6 = max(transaction_amount_usd),
              min_pur_amt_6 = min(transaction_amount_usd),
              avg_interval_pur_6 = avgInterval(transaction_date),
              max_interval_pur_6 = maxInterval(transaction_date),
              min_interval_pur_6 = minInterval(transaction_date),
              last_pur_time_6    = recentInterval(min(transaction_date),
                                                  max(transaction_date),
                                                  transaction_date)
    ) ->
  rollup
  
  ## Merge aggregated transaction data with demographic data.
  
  merged <- merge(rollup, demog, by=c("account_id")) %>% data.frame()
  
  merged
}

# Produce a prediction function that can use the model

creditRiskPrediction <- function(newdata)
{ 
  newdata <- rxImport(inData=newdata, stringsAsFactors=TRUE)
  pred <- rxPredict(modelObject=model_rxtrees, data=newdata)[, c(1, 3)]
  pred <- cbind(newdata$account_id, pred)
  names(pred) <- c("account_id", "predicted_default", "predicted_prob")
  pred 
}

# Test function locally by printing results

testtxn <- txn %>% 
  filter(accountID == "A1899946977632390") %>%
  data.frame() %T>%
  {head(.) %>% print()}

testdemog <- demog %>% 
  filter(accountID == "A1899946977632390") %>%
  data.frame() %T>%
  {head(.) %>% print()}

testdata <- creditRiskProcess(txn=testtxn, demog=testdemog)

print(testdata)

pred <- creditRiskPrediction(newdata=testdata)

print(pred)
```

### 2.2 Publish as web services

The second procedure is to publish the model as a web service by following the below steps.

Step 1: From your local R IDE, log into Microsoft R Server with your credentials using the appropriate authentication function from the `mrsdeploy` package (remoteLogin or remoteLoginAAD). 

For simplicity, the code below uses the basic local admin account for authentication with the remoteLogin function and `session = false` so that no remote R session is started. 

```{r, message=FALSE, warning=FALSE, error=FALSE}
# Use `remoteLogin` to authenticate with R Server using 
# the local admin account. Use session = false so no 
# remote R session started

remoteLogin("http://localhost:12800", 
         username="admin", 
         password="P@ssw0rd",
         session=FALSE)
```

Now, you are successfully connected to the remote R Server.

Step 2: Publish the model as a web service to R Server using the `publishService()` function from the `mrsdeploy` package. 

In this example, you publish two web services. One web service is called "crprocService" using the function `creditRiskProcess()`. As inputs, the service takes transaction records and demographic $ bank account information. As output, an R data frame having been aggregated, merged and cleansed is generated. Another web service is called "crpredService" using the model `model_rxtrees` and the function `creditRiskPrediction()`. As an input, the service takes an R data frame containing account-level transaction statistics and demographic & bank account information represented as numerical or categorical. As an output, an R data frame including the account id, the predicted label of default, and the probability of default for the given individual account, has of being achieved with the pre-defined credit risk prediction function. 

When publishing, you must specify, among other parameters, a service name and version, the R code, the inputs, as well as the outputs that application developers will need to integrate in their applications. 

```{r, message=FALSE, warning=FALSE, error=FALSE}
# Publish web services

apiCreditRiskProcess <- publishService(
                        "crprocService",
                        code=creditRiskProcess,
                        inputs=list(txn="data.frame", 
                                    demog="data.frame"),
                        outputs=list(merged="data.frame"),
                        v="v1.0.0")

apiCreditRiskPrediction <- publishService(
                           "crpredService",
                            code=creditRiskPrediction,
                            model=model_rxtrees,
                            inputs=list(newdata="data.frame"),
                            outputs=list(pred="data.frame"),
                            v="v1.0.0")
```

### 2.3 Test the services by consuming it in R

After publishing it , we can consume the service in R directly to verify that the results are as expected.

```{r, message=FALSE, warning=FALSE, error=TRUE}
# Consume service by calling function, `creditRiskProcess` contained in this service

resultProc <- apiCreditRiskProcess$creditRiskProcess(txn=testtxn, demog=testdemog)

# Print response output named 'merged'

print(resultProc$output("merged")) 

# Consume service by calling function, `creditRiskPrediction` contained in this service

testdata <- resultProc$output("merged")

testdata[! colnames(testdata) %in% c("account_id", "marital_status", "sex", "education", "account_setup_status", "bad_flag")] %<>%
  lapply(as.character) %>%
  lapply(as.numeric) %>%
  unlist()

resultPred <- apiCreditRiskPrediction$creditRiskPrediction(newdata=testdata)

# Print response output named `pred`

print(resultPred$output("pred")) 
```

### 2.4 Update the web service

In the process of production, we could manage and update the web service timely.

```{r, message=FALSE, warning=FALSE, error=TRUE}
# Load the pre-trained optimal model obtained from the template of CreditRiskScale.

load(file="model_rxtrees.RData")

model_rxtrees

# Update the web service api

apiCreditRiskPrediction <- updateService(name="crpredService", 
                                         v="v1.0.0",
                                         model=model_rxtrees,
                                         descr="Update the model hyper-parameters")

# Get service and assign service to the variable `api`.

apiCreditRiskPrediction <- getService("crpredService", "v1.0.0")

# Re-test the updated service by consuming it

resultPred <- apiCreditRiskPrediction$creditRiskPrediction(testdata)

# Print response output named `answer`

print(resultPred$output("pred")) 
```

### 2.5 Application Integration

Last but not least, we can get the json file that is needed for application integration.

```{r, message=FALSE, warning=FALSE, error=TRUE}
# Get this service's `swagger.json` file that is needed for web application integration

swagger1 <- apiCreditRiskProcess$swagger(json = TRUE)

cat(swagger1, file="swagger1.json", append=FALSE) 

swagger2 <- apiCreditRiskPrediction$swagger(json = TRUE)

cat(swagger2, file="swagger2.json", append=FALSE) 

# Delete the service to make the script re-runable

deleteService(name="crprocService", v="v1.0.0")
deleteService(name="crpredService", v="v1.0.0")
```
