---
title: "Solar power forecasting with Long Short-Term Memory (LSTM)"
author: "Le Zhang, Data Scientist, Cloud and AI, Microsoft"
date: '`r Sys.Date()`'
output:
  html_notebook: default
---

This accelerator is a reproduction of CNTK tutorial 106 B - using LSTM
for time series forecasting in R. The original tutorial can be found [here](https://github.com/Microsoft/CNTK/blob/master/Tutorials/CNTK_106B_LSTM_Timeseries_with_IOT_Data.ipynb).

The accelerator here mainly demonstrates how one can use `keras` R interface 
together with CNTK backend, to train a LSTM model for solar power forecasting,
in a Azure Data Science Virtual Machine (DSVM).

## 1 Introduction

### 1.1 Context.

[Solar power forecasting](https://en.wikipedia.org/wiki/Solar_power_forecasting)
is a challenging and important problem. Analyzing historical time-series data of
solar power generation may help predict the total amount of energy produced by 
solar panels. 

More discussion about solar power forecasting can be found in Wikipedia page. The
problem in this accelerator, is a simplified model, which is to merely illustrate
how an R based LSTM model can be trained in an Azure DSVM.

### 1.2 LSTM

LSTM is a type of Recurrent Neural Network, which is featued by its capability in
modeling long-term dependencies. It's been pratically applied in many fields such
as Natural Language Processing (NLP), action recognition, time series prediction,
etc. 

While a comprehensive discussion of LSTM is not the focus of this accelerator, 
more information can be found in [Chris Olah's blog](http://colah.github.io/posts/2015-08-Understanding-LSTMs/).

### 1.3 Cognitive Toolkit and keras

#### 1.3.1 Cognitive Toolkit

[Microsoft Cognitive Toolkit (previously known as CNTK)](https://www.microsoft.com/en-us/cognitive-toolkit/) is a free, easy-to-use,
open-source, and commercial-grade toolkit that trains deep learning algorithms
to learn like the human brains. 

It is featured by 
* Highly optimized and built-in components that handle multi-dimensional data 
from different language environment, deal with various types of deep learning
algorithms, add user-defined core components on the GPU, etc.
* Efficience resource usage that avails parallelism with multiple GPU/machines.
* Easy expression of neural network with full APIs of Python, C++, and BrainScript.
* Training support on Azure.

#### 1.3.2 Keras

[Keras](https://keras.io/) is a high-level neural networks API, that is capable of running various 
backends such as Cognitive Toolkit, Tensorflow, and Theano. It made experimenting
deep learning neural networks from idea to result easier than ever before.

#### 1.3.3 Cognitive Toolkit + Keras in R

Since version 2.0, Cognitive Toolkit starts to support keras. 

Cognitive Toolkit has not supported R yet. However, by using [Keras R interface](https://rstudio.github.io/keras/), one can try training neural network
models by using keras API with Cognitive Toolkit backend.

## 2 Cloud resource deployment

Azure cloud platform offers varieties of resources for elastically running 
scalable analytical jobs. Especially, VMs or VM clusters incorporated with 
high-performance computing engines make it convenient for researchers and 
developers to prototype and validate models easily.

The following sections demonstrate how to train an LSTM model on a DSVM with 
Cognitive Toolkit and Keras in R.

NOTE: **the script for demonstrating Cognitive Toolkit + Keras can also be run
in a local environment, but one needs to manually download and install
Cognitive Toolkit and its dependencies, Keras, Keras R package and its
dependencies, CUDA Toolkit (if GPU device is available and GPU acceleration 
is wanted).**

### 2.1 Data Science Virtual Machine (DSVM)

[Azure DSVM](https://docs.microsoft.com/en-us/azure/machine-learning/machine-learning-data-science-provision-vm) is a curated VM that is pre-installed with 
rich set of commonly used data science and AI development tools such as R/Python
environment, Cognitive Toolkit, SQL Server, etc.

DSVM is a desirable workplace to experiment, prototype, and productize data 
analytical and AI solutions. Elasticity of the offering also guarantee the cost
effectiveness, which makes it more economically efficient compared to on-premise
servers.

### 2.2 Configuration and setup

Both Cognitive Toolkit and keras are pre-installed on DSVM. However, so far the 
keras R packages as well as its dependencies are not available. Preliminary 
installation and configuration are therefore required.

### 2.3 Resource deployment with `AzureDSVM`

[`AzureDSVM`](https://github.com/Azure/AzureDSVM) is an R packages that allow R users to directly interact with Azure
account to administrate DSVM instances.

To fire up a DSVM, one just needs to specify information such as DSVM name, user
name, operating system, VM size, etc. For example, the following script fires up
a Ubuntu DSVM of size D1_V2, located at Southeast Asia.

```{r}
# load the packages

library(AzureSMR)
library(AzureDSVM)
```

```{r}
# Credentials for authentication against Azure account are preserved in a json
# formmated file, named "config.json", which is located at ~/.azuresmr

# The credentials needed for authetication include Client ID, Tenant ID, 
# authentication key, password, and public key.

settingsfile <- getOption("AzureSMR.config")
config <- read.AzureSMR.config()
```

```{r}
# Authentication with the credential information. 

asc <- createAzureContext()

with(config,
     setAzureContext(asc, tenantID=tenantID, clientID=clientID, authKey=authKey)
)
azureAuthenticate(asc)
```

```{r}
# location and resource group name.

dsvm_location <- "southeastasia"
dsvm_rg       <- paste0("rg", paste(sample(letters, 3), collapse=""))

# VM size, operating system, and VM name.

dsvm_size     <- "Standard_D1_v2"
dsvm_os       <- "Ubuntu"
dsvm_name     <- paste0("dsvm", 
                        paste(sample(letters, 3), collapse=""))

# VM user name, authentication method (password in this case), and login password.

dsvm_username <- "dsvmuser"
dsvm_authen   <- "Password"
dsvm_password <- config$PASSWORD
```

```{r eval=FALSE}
# deploy the DSVM.

deployDSVM(asc, 
           resource.group=dsvm_rg,
           location=dsvm_location,
           hostname=dsvm_name,
           username=dsvm_username,
           size=dsvm_size,
           os=dsvm_os,
           authen=dsvm_authen,
           password=dsvm_password,
           mode="Sync")
```

Note since originally a DSVM does not have keras R interface installed and 
configured. A post-deployment installation and configuration of the package can 
be achieved by adding an extension to the deployed DSVM. Basically it runs a 
shell script that is located on a remote place.

```{r}
# URL of the shell script and the command to run the script.

dsvm_fileurl <- "https://raw.githubusercontent.com/yueguoguo/Azure-R-Interface/master/demos/demo-5/script.sh"
dsvm_command <- "sudo sh script.sh"
```

```{r eval=FALSE}
# Add extension to the DSVM.

addExtensionDSVM(asc,
                 location=dsvm_location,
                 resource.group=dsvm_rg,
                 hostname=dsvm_name,
                 os=dsvm_os, 
                 fileurl=dsvm_fileurl, 
                 command=dsvm_command)
```

### 2.4 Remote access to the DSVM.

After a successful deployment and extension, the DSVM can be remotely accessed 
by 
1. Rstudio Server - http://<dsvm_name>.<location>.cloudapp.azure.com:8787
2. Jupyter Notebook - https://<dsvm_name>.<location>.cloudapp.azure.com:8000
3. X2Go client.

NOTE: **it was found that keras R interface does not work well in Rstudio server
owing to SSL certificate issue.**

Idealy in the R session of the remote DSVM, typing the following

```{r eval=FALSE}
library(keras)

backend()
```

will show the message of "Using CNTK backend...", which means the interface can 
detect and load Cognitive Toolkit backend.

## 2.5 Model building

After all the set up, the model can be created by using Keras R interface 
functions. As the model building follows the original CNTK tutorial, text for 
introduction and description will not be duplicated here.

Script of the whole step-by-step tutorial is available [here](../Code/lstm.R)

### 2.6 Run script on the DSVM

The script can be tried on the deployed DSVM in various ways.

1. Jupyter Notebook - access the Jupyter Hub hosted by the DSVM via https://<dsvm_name>.<location>.cloudapp.azure.com:8000. Create an R-kernel 
notebook to run the script.
2. X2Go client - create a new X2Go session for remote desktop of that machine.
The script can be copied onto the DSVM via either SSH or any SSH-based file
transfer software, and then be run in Rstudio desktop version.

NOTE: **it was found that if the script is run with R console or Rscript in 
command line, GPU device will not be activated for acceleration. So running
the script in Rstudio IDE is recommended.**

## 3 Closing

After the experiment, it is recommended to either stop and deallocate, or 
destroy the computing resource, if it is not needed.

```{r eval=FALSE}
# Stop and deallocate the DSVM.

operateDSVM(asc, dsvm_rg, dsvm_name, "Stop")
```

```{r eval=FALSE}
# Delete the resource group.

azureDeleteResourceGroup(asc, dsvm_rg)
```
